---
title: "Darwin Core mapping"
subtitle: "For: Checklist of the belgian Formicidae"
author:
- Lien Reyserhove
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document describes how we map the checklist data to Darwin Core. The source file for this document can be found [here](https://github.com/trias-project/ad-hoc-checklist/blob/master/src/dwc_mapping.Rmd).

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries:

```{r}
library(tidyverse)      # To do data science
library(magrittr)       # To use %<>% pipes
library(here)           # To find files
library(janitor)        # To clean input data
library(digest)         # To generate hashes
```

# Read source data

The data is maintained in [this Google Spreadsheet](https://docs.google.com/spreadsheets/d/1Rvb9qwpSrflZ9bABvGa-F7BhHROsQCSn/edit#gid=1231119869).

Read the relevant worksheet (published as csv):

```{r read_source_data}
# input_data <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTDe07wnCsBlbTnY7-5loI8Bd5Z4G_ydBdi0MhgMUz23CiAAxzSf0KSi3UL6y2ePA/pub?gid=1231119869&single=true&output=csv")

input_data <- read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vTl8IEk2fProQorMu5xKQPdMXl3OQp-c0f6eBXitv0BiVFZ3JSJCde0PtbFXuETgguf6vK8b43FDX1C/pub?gid=1172557183&single=true&output=csv")
```

Copy the source data to the repository to keep track of changes:

```{r}
write_csv(input_data, here("data", "raw", "belgian_formicidae_dump.csv"), na = "")
```

Preview data: 

```{r}
input_data %>% head()
```
# Preprocessing

## Tidy data

Clean data somewhat:

```{r}
input_data %<>%
  remove_empty("rows") %>%    # Remove empty rows
  clean_names()               # Have sensible (lowercase) column names
```

## Scientific names

No cleaning required.

## Taxon IDs

To link taxa with information in the extension(s), each taxon needs a unique and relatively stable `taxonID`. Here we create one in the form of `dataset_shortname:taxon:hash`, where `hash` is unique code based on scientific name and kingdom (that will remain the same as long as scientific name and kingdom remain the same):

```{r}
vdigest <- Vectorize(digest) # Vectorize digest function to work with vectors
input_data %<>% mutate(taxon_id = paste(
  "belgian-formicidae-checklist",
  "taxon",
  vdigest(paste(scientific_name, kingdom), algo = "md5"),
  sep = ":"
))
```

Preview data:

```{r}
input_data %>% head()
```

# Darwin Core mapping

## Taxon core

Create a dataframe with unique taxa only (ignoring multiple distribution rows):

```{r}
taxon <- input_data %>% distinct(taxon_id, .keep_all = TRUE)
```

Map the data to [Darwin Core Taxon](http://rs.gbif.org/core/dwc_taxon_2015-04-24.xml).
 
### language

```{r}
taxon %<>% mutate(dwc_language = "en")
```

### license

```{r}
taxon %<>% mutate(dwc_license = "http://creativecommons.org/publicdomain/zero/1.0/")
```

### rightsHolder

```{r}
taxon %<>% mutate(dwc_rightsHolder = "RBINS") 
```

### accessRights

```{r}
taxon %<>% mutate(dwc_accessRights = "https://www.inbo.be/en/norms-data-use") 
```

### datasetID

```{r}
taxon %<>% mutate(dwc_datasetID = "https://doi.org/10.15468/ub8mcm") 
```

### institutionCode

```{r}
taxon %<>% mutate(dwc_institutionCode = "RBINS") 
```

### datasetName

```{r}
taxon %<>% mutate(dwc_datasetName = "Checklist of the ants in Belgium") 
```

### taxonID

```{r}
taxon %<>% mutate(dwc_taxonID = taxon_id)
```

### scientificName

```{r}
taxon %<>% mutate(dwc_scientificName = scientific_name) 
```

### kingdom

```{r}
taxon %<>% mutate(dwc_kingdom = kingdom)
```

### phylum

```{r}
taxon %<>% mutate(dwc_phylum = phylum)
```

### class

```{r}
taxon %<>% mutate(dwc_class = class)
```

### order

```{r}
taxon %<>% mutate(dwc_order = order)
```

### family

```{r}
taxon %<>% mutate(dwc_family = family)
```

### genus

```{r}
taxon %<>% mutate(dwc_genus = genus)
```

### taxonRank

```{r}
taxon %<>% mutate(dwc_taxonRank = taxon_rank)
```

### nomenclaturalCode

```{r}
taxon %<>% mutate(dwc_nomenclaturalCode = nomenclatural_code)
```

## Distribution extension

Create a dataframe with all data (including multiple distributions):

```{r}
distribution <- input_data
```

Map the data to [Species Distribution](http://rs.gbif.org/extension/gbif/1.0/distribution.xml).

### taxonID

```{r}
distribution %<>% mutate(dwc_taxonID = taxon_id) 
```

### locationID

Currently map for Belgian regions only:

```{r}
distribution %<>% mutate(dwc_locationID = case_when(
  is.na(location) & country_code == "BE" ~ "ISO_3166-2:BE",
  location == "Flanders" ~ "ISO_3166-2:BE-VLG",
  location == "Wallonia" ~ "ISO_3166-2:BE-WAL",
  location == "Brussels" ~ "ISO_3166-2:BE-BRU"
))
```

### locality

Currently map for Belgian regions only:

```{r}
distribution %<>% mutate(dwc_locality = case_when(
  is.na(location) & country_code == "BE" ~ "Belgium",
  location == "Flanders" ~ "Flemish Region",
  location == "Wallonia" ~ "Walloon Region",
  location == "Brussels" ~ "Brussels-Capital Region"
))
```

### countryCode

```{r}
distribution %<>% mutate(dwc_countryCode = country_code)
```

### occurrenceStatus

```{r}
distribution %<>% mutate(dwc_occurrenceStatus = occurrence_status)
```

### establishmentMeans

```{r}
distribution %<>% mutate(dwc_establishmentMeans = case_when(
  degree_of_establishment == "established" | 
  degree_of_establishment == "casual" |
  degree_of_establishment == "invasive" ~ "introduced",
  TRUE ~ degree_of_establishment))
```

### degreeOfEstablishment

```{r}
distribution %<>% mutate(dwc_degreeOfEstablishment = degree_of_establishment) 
```

### pathway

```{r}
distribution %<>% mutate(dwc_pathway = pathway_lien) 
```


### eventDate

```{r}
distribution %<>% mutate(dwc_eventDate = case_when(
  is.na(date_first_observation) & is.na(date_last_observation) ~ "",
  is.na(date_first_observation) & !is.na(date_last_observation) ~ as.character(date_last_observation),
  !is.na(date_first_observation) & is.na(date_last_observation) ~ as.character(date_first_observation),
  !is.na(date_first_observation) & !is.na(date_last_observation) ~ as.character(paste(date_first_observation, date_last_observation, sep = "/")))) 
```

### threatStatus

### source

```{r}
distribution %<>% mutate(dwc_source = source)
```

### occurrenceRemarks

```{r}
distribution %<>% mutate(dwc_occurrenceRemarks = occurrence_remarks) 
```

## Species profile extension

In this extension we will express broad habitat characteristics of the species (e.g. `isTerrestrial`) from `realm`.

Create a dataframe with unique taxa only (ignoring multiple distribution rows):

```{r}
species_profile <- input_data %>% distinct(taxon_id, .keep_all = TRUE)
```

Only keep records for which `realm` is not empty:

```{r}
species_profile %<>% filter(!is.na(realm))
```

Inspect values:

```{r}
species_profile %>%
  group_by(realm) %>%
  count()
```

Map the data to [Species Profile](http://rs.gbif.org/extension/gbif/1.0/speciesprofile.xml).

### taxonID

```{r}
species_profile %<>% mutate(dwc_taxonID = taxon_id)
```

### isMarine, isFreshwater, isTerrestrial

```{r}
species_profile <- species_profile %>% 
  mutate(dwc_isMarine = "FALSE") %>% 
  mutate(dwc_isFreshwater = "FALSE") %>%
  mutate(dwc_isTerrestrial = "TRUE") 
```


## Description extension

In the description extension we want to include several important characteristics (hereafter referred to as descriptors) about the species:
- Native range

We will express native range information from `native_range`, separated and gathered.

Create a dataframe with unique taxa only (ignoring multiple distribution rows):

```{r}
description <- input_data %>% distinct(taxon_id, .keep_all = TRUE)
```

Separate values on `|` in a maximum of 3 columns:

```{r}
description %<>% separate(
  native_range,
  into = paste0("range_", c(1:4)),
  sep = " \\| ",
  extra = "drop"
)
```
Gather and trim values:

```{r}
description %<>% gather(key, value, starts_with("range_"), na.rm = TRUE) %>%
  mutate(value = str_trim(value))
```

Inspect values:

```{r}
description %>%
  group_by(value) %>%
  count()
```

Clean native range information in `value` somewhat:

```{r}
description %<>% mutate(value = str_to_title(value))
```

### taxonID

```{r}
description %<>% mutate(dwc_taxonID = taxon_id)
```

### description

```{r}
description %<>% mutate(dwc_description = value)
```

### type

```{r}
description %<>% mutate(dwc_type = "native range")
```

### language

```{r}
description %<>% mutate(dwc_language = "en")
```

# Post-processing

Only keep the Darwin Core columns:

```{r}
taxon %<>% select(starts_with("dwc_"))
distribution %<>% select(starts_with("dwc_"))
species_profile %<>% select(starts_with("dwc_"))
description %<>% select(starts_with("dwc_"))
```

Drop the `dwc_` prefix:

```{r}
colnames(taxon) <- str_remove(colnames(taxon), "dwc_")
colnames(distribution) <- str_remove(colnames(distribution), "dwc_")
colnames(species_profile) <- str_remove(colnames(species_profile), "dwc_")
colnames(description) <- str_remove(colnames(description), "dwc_")
```

Sort on `taxonID` (to maintain some consistency between updates of the dataset):

```{r}
taxon %<>% arrange(taxonID)
distribution %<>% arrange(taxonID)
species_profile %<>% arrange(taxonID)
description %<>% arrange(taxonID)
```

Preview taxon core:

```{r}
taxon %>% head()
```

Preview distribution extension:

```{r}
distribution %>% head()
```

Preview species profile extension:

```{r}
species_profile %>% head()
```

Preview description extension:

```{r}
description %>% head(10)
```

Save to CSV:

```{r}
write_csv(taxon, here("data", "processed", "taxon.csv"), na = "")
write_csv(distribution, here("data", "processed", "distribution.csv"), na = "")
write_csv(species_profile, here("data", "processed", "speciesprofile.csv"), na = "")
write_csv(description, here("data", "processed", "description.csv"), na = "")
```
